SmartCheck

### 2012-12-03 paper/paper.tex
paper updates.

\usepackage{pgf}
%% \usepackage{tikz}
%% \usetikzlibrary{arrows,automata}

@@

@@

The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
aspect of the Haskell development cycle.

Because QuickCheck generates
random values for testing, counterexamples it finds may not be substantially
larger than a minimal counterexample.  In their original QuickCheck
paper~\cite{qc}, the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counter examples found are very large and it is difficult to go
back to the property and understand why it is a counter example.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that defines a function
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a] 
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

-- Property
prop :: B -> Property
prop p = pre p ==> post p
\end{code}
  \caption{QuickCheck and SmallCheck for a product type input.}
  \label{fig:initial}
\end{figure}

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.  Indeed, the principal motivation for one
of Haskell's first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{E.g., see the user question posed on StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields of
\ttp{B} is less than 64.  \ttp{prop} seems reasonable at first glance, until
one realizes that due to underflow, the property can be violated.  For example,
consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
a value contains 70 \ttp{Int16} values!

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided, which also
rely on QuickCheck's default instances for the list type, the list of potential
counterexamples generated by \ttp{shrink} can contain more than $10^{10}$
elements.  This is an enormous state space to search for a smaller example!

@@

Typically, a user tries a brute-force approach to reduce the number of potential
counterexamples, for example by truncating lists using the \ttp{take n} function
that returns the first $n$ elements of a list.  The trade-off is quicker
shrinking with a lower probability of finding a smaller counterexample.  For
example, redefining shrink for type \ttp{B} as follows
%
\begin{code}
shrink (B a b c d) = [ B w x y z | w <- tk a, x <- tk b
                                 , y <- tk c, z <- tk d ]
  where tk x = take 10 (shrink x)
\end{code}
%
\noindent
controls the exponential blowup of the state-space by taking the first ten
elements of each list, so the cross product is limited to 1000 elements.  The
downside is that potentially smaller counterexamples may be omitted.

What if we omit the need for shrinking counterexamples altogether?  SmallCheck
is another testing framework for Haskell.  SmallCheck is guaranteed to return
the a smallest counterexample, if one exists~\cite{sc}.  SmallCheck does this by
enumerating all possible inputs, ordered from smallest to largest, up to some
user-defined bound.  While SmallCheck is effective for testing many programs and
properties (in accordance with the \emph{small scope hypothesis}~\cite{jackson}),
counterexamples to even relatively simple properties may be infeasible to
discover due to state-space explosion.

@@

Finally, even if we could overcome the problems described above with QuickCheck
and SmallCheck, we are left with two others.  First, when a
counterexample is discovered, one might wish to search for different
counterexamples.  There currently is no automated way to add a predicate
characterizing a counterexample as preconditions to a property to strengthen the
property.  Second, and more generally, a counterexample is usually
representative of other counterexamples, but that generalization is left to the
user to discover for herself.  Really, what would be useful would be not just a
minimal counterexample, but a quantified formula characterizing a set of
counterexamples.  Such a formula has the added benefit of replacing a
potentially large sub-term of a composite data type value with a quantifier.

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples.  To motivate the benefit of SmartCheck, consider the
table in Figure~\ref{fig:results}.  The table characterizes testing the program
shown in Figure~\ref{fig:initial} that we have described so far.  Four runs are
presented.  In the first three, we use QuickCheck to generate counterexamples.
The first column contains the results of using QuickCheck without shrinking to
provide a baseline.  The next two columns show the result of using QuickCheck
with definitions of \ttp{shrink} that limit the lists of fields of \ttp{B} to 10
(the ``QC (10)'' tests) and 20 (the ``QC (20)'' tests) elements, respectively,
using a definition like the truncated implementation of \ttp{shrink} for the
type \ttp{B} discussed earlier.  In each row, we show the maximum, mean, and
median number of seconds required to by QuickCheck to generate a counterexample
over 100 runs.  In the last row, we show the mean size of the final value
returned, measured by the taking the cumulative length of the lists of the
fields of \ttp{B}.  Notice that while typically counterexamples are returned in
a fraction of a second, there are outliers taking 21 seconds and 125 seconds for
the two tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

In the last column are the results under the same condition for SmartCheck.  Not
only does it execute faster than QuickCheck with shrinking, it has significantly
smaller outliers.  The most striking difference, however, is the size of the
counterexample returned, reducing on average the size of a counterexample found
by QuickCheck without shrinking by a factor of 14.

The purpose of this paper is to explain first how to efficiently and generically
generate small counterexamples for algebraic data types and second how to
generalize the counterexamples..

@@

\item Using generic programming techniques, we describe efficient counterexample
  reduction strategies for algebraic data types.

\item We describe an approach to counterexample generalization, to present the
  user with a formula describing a set of counterexamples to a property.

\item We present an approach to generically strengthen a property with a
  precondition that characterizes an already-discovered counterexample for the
  purpose of finding new, dissimilar counterexamples.

\item More generally, this paper is the first to explore the idea of efficient,
  generic counterexample reduction and generalization for functional program testing.
\end{itemize}

@@

\newpage

\todo{insight: shrink based on arbitrary}

\section{Shrinking Data}

>>>>

\usepackage{pgf}

\usepackage{listings}
\lstset{ %
  language=Haskell,                 % the language of the code
  basicstyle=\scriptsize,           % the size of the fonts that are used
  %% for the code
  %%firstnumber=0, % index to start numbering at
  %%numbers=left,                      % where to put the line-numbers
  %% numberstyle=\tiny\color{gray},  % the style that is used for the
  %% line-numbers
  stepnumber=5,                   % the step between two line-numbers. If it's
                                  % 1, each line 
  %%                              % will be numbered
  numbersep=5pt,                  % how far the line-numbers are from the code
  xleftmargin=0ex,
  %% backgroundcolor=\color{white},      % choose the background color. You must
  %% add \usepackage{color}
  %% showspaces=false,               % show spaces adding particular underscores
  %% showstringspaces=false,         % underline spaces within strings
  %% showtabs=false,                 % show tabs within strings adding
  %% particular underscores
  %% frame=single,                   % adds a frame around the code
  %% rulecolor=\color{black},        % if not set, the frame-color may be
  %% changed on line-breaks within not-black text (e.g. commens (green here))
  %% tabsize=2,                      % sets default tabsize to 2 spaces
  %% captionpos=b,                   % sets the caption-position to bottom
  %% breaklines=true,                % sets automatic line breaking
  %% breakatwhitespace=false,        % sets if automatic breaks should only
  %% happen at whitespace
  %% title=\lstname,                   % show the filename of files included
  %% with \lstinputlisting;
  %%                                 % also try caption instead of title
  %% keywordstyle=\color{blue},          % keyword style
  %% commentstyle=\color{dkgreen},       % comment style
  %% stringstyle=\color{mauve},          % string literal style
  escapeinside={@*}{*@},            % if you want to add a comment within your
                                    % code
  %% morekeywords={*,...}           % if you want to add more keywords to
  %% the set
}

%% \usepackage{tikz}
%% \usetikzlibrary{arrows,automata}

@@

\newcommand{\sub}[1]{\(\sb{#1}\)}

@@

The QuickCheck testing framework was a revolutionary step-forward in
type-directed random testing~\cite{qc}.  Nowadays, QuickCheck is a standard
aspect of the Haskell development cycle.  Because QuickCheck generates random
values for testing, counterexamples it finds may not be substantially larger
than a minimal counterexample.  In their original QuickCheck paper~\cite{qc},
the authors report the following user experience by Andy Gill:
%
\begin{quote}
Sometimes the counterexamples found are very large and it is difficult to go
back to the property and understand why it is a counterexample.
\end{quote}
%
\noindent
QuickCheck defines a type class \ttp{Arbitrary} that presents a method
\ttp{arbitrary} for generating random values of a given type.  Gill added another
method to the type class:
%
\begin{code}
smaller :: a -> [a] 
\end{code}
%
\noindent
The purpose of \ttp{smaller} is to generate strictly smaller values (according to
some measure) from a given counterexample to attempt to find a smaller
counterexample.  Today, \ttp{smaller} is called \ttp{shrink}.

@@

-- Property
prop :: B -> Property
prop p = pre p ==> post p
\end{code}
  \caption{QuickCheck and SmallCheck for a product-type input.}
  \label{fig:initial}
\end{figure}

While \ttp{shrink} is user-defined and can have any type-correct implementation,
it is typically defined using structural recursion for composite data types.
For example, consider the \ttp{shrink} instances given to data types \ttp{A} and
\ttp{B} in Figure~\ref{fig:initial}.  Indeed, because of the regular structure
of \ttp{shrink} implementations, the principal motivation for one
of Haskell's first generics package was for providing generic definitions for
\ttp{shrink}~\cite{syb}.

However, this straightforward approach to defining \ttp{shrink} instances may
either miss minimal counterexamples, be too inefficient to be practical, or
both.\footnote{E.g., for a recent real-world case, see the user question posed on StackOverflow:
  \url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}.}
For example, consider again Figure~\ref{fig:initial}.  The data type \ttp{B} is
a product type over lists of 16-bit signed integers, wrapped by a constructor.
Consider the property \ttp{prop} in the figure, stating that if the sum of each
list in a field of \ttp{B} is less than 16, then the sum of the four fields of
\ttp{B} is less than 64.  \ttp{prop} seems reasonable at first glance, until
one realizes that due to underflow, the property can be violated.  For example,
consider the value
%
\begin{code}
B [A (-32769)] [] [] []
\end{code}
%
\noindent
Without shrinking, QuickCheck finds large examples to the property; on average,
QuickCheck returns a value contains 70 \ttp{Int16} values!

Thus, it pays to define \ttp{shrink}.  Unfortunately, even for a simple property
and program like this one, the definition for \ttp{shrink} given in
Figure~\ref{fig:initial} produces an intractable number of potential
counterexamples to test; with the definitions and property provided, which also
rely on QuickCheck's default instances for the list type, the list of potential
counterexamples generated by \ttp{shrink} can contain more than $10^{10}$
elements.  This is an enormous state space to search for a smaller example!  In
particular, if a counterexample is not found does not finish in a reasonable
amount of time.

@@

Typically, a user tries a brute-force approach to reduce the number of potential
counterexamples, for example, by truncating lists using the \ttp{take n} function
that returns the first $n$ elements of a list.  The trade-off is quicker
shrinking with a lower probability of finding a smaller counterexample.  For
example, redefining shrink for type \ttp{B} as follows
%
\lstset{ caption={Truncated shrinking function for data type \ttp{B}.}
       , label=lst:newshrink
       ,captionpos=b }
\begin{lstlisting}
len = 10

shrink (B a b c d) = [ B w x y z | w <- tk a, x <- tk b
                                 , y <- tk c, z <- tk d ]
  where tk x = take len (shrink x)
\end{lstlisting}
%% \begin{code}
%% \end{code}
%
\noindent
controls the exponential blowup of the state-space by taking the first ten
elements of each list, so the cross product is limited to 1000 elements.  The
downside is that potentially smaller counterexamples may be omitted.

The programmer might ask herself: ``What if we omit the need for shrinking
counterexamples altogether?''  SmallCheck is another testing framework for
Haskell that does just this.  SmallCheck is guaranteed to return the a smallest
counterexample, if one exists~\cite{sc}.  SmallCheck does this by enumerating
all possible inputs, ordered from smallest to largest, up to some user-defined
bound.  While SmallCheck is effective for testing many programs and properties
(in accordance with the \emph{small scope hypothesis}~\cite{jackson}),
counterexamples to even relatively simple properties may be infeasible to
discover due to state-space explosion.

@@

Finally, even if we could overcome the problems described above with QuickCheck
and SmallCheck, we are left with two others.  
\begin{itemize}
  \item First, when a counterexample is discovered to some property, say with
    type
%
\begin{code}
prop :: B -> Bool      
\end{code}
%
\noindent
    one might wish to search for different counterexamples to \ttp{prop}.
    Ideally, the testing framework would automatically add a predicate
    characterizing (and perhaps generalizing) a counterexample 
%
    \begin{code}
cex :: B -> Bool      
    \end{code}
%
\noindent as a precondition to \ttp{prop} to define a new property
\begin{code}
prop' b = not (cex b) ==> prop b
\end{code}
%
\noindent
where \ttp{==>} is implication to generate new counterexamples to the same property.

  \item Second, and more generally, a counterexample is usually representative
    of other counterexamples, but that generalization is left to the user to
    discover for herself.  Really, what would be useful would be not just a
    minimal counterexample, but a quantified formula characterizing a set of
    counterexamples.  Such a formula has the added benefit of replacing a
    potentially large sub-term of a composite data type value with a quantifier.
\end{itemize}

@@

\paragraph{SmartCheck}
Motivated by these limitations of QuickCheck and SmallCheck, we have developed
\emph{SmartCheck}.  SmartCheck extends QuickCheck to efficiently and generically
find small counterexamples as well as generalize them as described in the
preceding paragraph.  To motivate the benefit of SmartCheck, consider the table
in Figure~\ref{fig:results}.  The table characterizes testing the program shown
in Figure~\ref{fig:initial} that we have described so far.  Four trials are
presented.  In the first three, we use QuickCheck to generate counterexamples.
The first column contains the results of using QuickCheck without shrinking to
provide a baseline.  The next two columns show the result of using QuickCheck
with definitions of \ttp{shrink} that limit the lists of fields of \ttp{B} to 10
(Listing~\ref{lst:newshrink}) and 20 elements (Listing~\ref{lst:newshrink}, with
\ttp{len = 20}), respectively.  (Recall that with the original definition for
shrink from Figure~\ref{fig:initial}, some counterexamples cause QuickCheck to
not return after hours of computation.)  In each row, we show the maximum, mean,
and median number of seconds required by QuickCheck to generate a counterexample
over 100 runs.  In the last row, we show the mean size of the final value
returned, measured by the taking the cumulative length of the lists of the
fields of \ttp{B}.  Notice that while typically counterexamples are returned in
a fraction of a second, there are outliers taking 21 seconds and 125 seconds for
the two tests.  Both tests roughly cut the size of the counterexample in half;
increasing the number of potential counterexamples can result in a significant
performance penalty for this example without much benefit.

In the last column are the results for SmartCheck.  Not only does it execute
faster than QuickCheck with shrinking, it has significantly smaller outliers.
The most striking difference, however, is the size of the counterexample
returned, reducing on average the size of a counterexample found by QuickCheck
without shrinking by a factor of 14.

The purpose of this paper is to explain first how to efficiently and generically
generate small counterexamples for algebraic data types and second how to
generalize the counterexamples.

@@

\item an efficient counterexample reduction strategy for algebraic
  data types;

\item an approach to counterexample generalization, to present the
  user with a formula describing a set of counterexamples to a property;

\item an approach to generically strengthen a property with a precondition that
  characterizes an already-discovered counterexample for the purpose of finding
  new, dissimilar counterexamples;

\item an implementation based on generic programming that is not only more
  efficient that hand-written \ttp{shrink} instances in many cases but that
  generates the instances automatically for the user.

%% \item More generally, this paper is the first to explore the idea of
%%   efficient, generic counterexample reduction and generalization for
%%   functional program testing.
\end{itemize}

%% \noindent Moreover, the inability to quickly uncover understandable
%% counterexamples is a real-world problem with QuickCheck (or other functional
%% languages).  It works great when your tests pass,

@@

\section{Shrinking Data}
In this section, we describe how to efficiently shrink data values.  Before we
begin, recall the basic approach implemented by the \ttp{shrink} method:
generate a list of values, each of which is smaller than the current
counterexample.  Each of the new values generated has nothing in common with the
original counterexample, since each one is freshly generated by recursive
size-constrained calls to \ttp{arbitrary}.

Consider the problem of finding a small counterexample as a graph search
problem.  The set of vertexes are all possible well-typed values (the same as
the counterexample).  There is an edge from vertex $v$ to $v'$ if $v'$ is
smaller (by some measure).  The goal is to fine a vertex from the set of
smallest counterexamples in the graph.

The approach taken by \ttp{shrink} is the most naive graph search algorithm:
it simply generates a random set of potential vertexes to check in the graph.
However, it does not exploit that the original counterexample \ttp{cex} is
already in a neighborhood of likely counterexamples.  So rather than choose
random vertexes, it may be more efficient to test other values that are closely
related to \ttp{cex}.  By restricting the search to \ttp{cex}'s neighborhood, a
smaller counterexample might be missed, and only a local minimum is returned.
However, with a large state space, finding a global minimum by random sampling
vertexes in the graph is possible but too inefficient, as we saw from the
example in the introduction.

\paragraph{Smart reduction algorithm}
The algorithm we present for efficiently searching for new counterexamples can
be thought of as a recursive breadth-first traversal over the structure of the
current counterexample value, where the value is represented as a tree.  At each
node, during the traversal, we generate arbitrary \emph{structurally} smaller
sub-values and build a new value from that, leaving the remainder of the tree
unchanged.  Intuitively, by a structurally smaller value, we mean one with fewer
constructors.  We continue until we reach a fixed-point.

To present the algorithm, consider the \ttp{SubTypes} type class below for which
QuickCheck's \ttp{Arbitrary} class is a super-class.\footnote{This presentation
  deviates from the implementation; for example, we elide details regarding
  error handling.}
%
\begin{code}
class Arbitrary a => SubTypes a where
  size    :: a -> Int
  index   :: a -> Int -> Maybe SubVal
  replace :: a -> Int -> SubVal -> a
\end{code}
%
\noindent
\ttp{SubTypes} has three methods.  The \ttp{size} method returns the size of a
value, \ttp{index} returns a sub-value at an breadth-first index in a value, and
\ttp{replace} replaces a sub-value at a particular focus.  If the index is
out-of-bounds (too large or negative), \ttp{index} return \ttp{Nothing}.  The
method \ttp{replace} returns the original value for an out-of-bounds index or an
incorrectly-typed replacement value.  The \ttp{SubVal} type is an existential
data type, used as a generic container for sub-values from a counterexample.
%
\begin{code}
data SubVal = forall a. (Arbitrary a, SubTypes a) => SubVal a
\end{code}
%
\noindent
For example, consider a binary tree type
%
\begin{code}
data Tree = L | B Tree Tree
\end{code}
% 
\noindent
and the value \ttp{tree}, labeled with the indexes of its sub-values
%
\begin{code}
tree = B\sub{0} (B\sub{1} L\sub{3} 
             (B\sub{4} L\sub{6} L\sub{8}))
          (B\sub{2} L\sub{5} L\sub{7})
\end{code}
%
\noindent
Here are example applications of \ttp{SubTypes}; in the following, we show
the indexes with respect to the \ttp{tree} value:
%
\begin{code}
size tree = 9

index tree 4 = (Just . SubVal) (B\sub{4} L\sub{6} L\sub{8})

index tree 12 = Nothing

replace tree 2 (SubVal L) = 
  B\sub{0} (B\sub{1} L\sub{3} 
        (B\sub{4} L\sub{6} L\sub{8}))
     L\sub{2}
\end{code}
%
\noindent
We will sometimes refer to unwrapped value returned by \ttp{index a i} as the
\emph{$i$th sub-value of $a$}, so for example,

\begin{figure}[ht]
  \lstset{caption={}
%         ,xleftmargin=5ex,firstnumber=0,numbers=left
         }
  \begin{lstlisting}
reduce :: SubTypes a => a -> Property a -> IO a
reduce cex prop = reduce' 1
  where
  reduce' idx =
    case index cex idx of 
      Nothing -> return cex
      Just v  -> 
        case test cex idx v prop of
          Nothing   -> reduce' (idx + 1)
          Just cex' -> reduce cex' prop

test :: SubTypes a 
     => a -> Int -> SubVal -> Property a -> Maybe a
test cex idx v prop = go maxTries
  where
  go 0 = Nothing
  go n = let v'   = newVal v in
         let cex' = replace cex idx v' in
         if prop cex' then go (n - 1) else Just cex'

  newVal (SubVal a) = sizedArbitrary a (size a)
  \end{lstlisting}
  \caption{Counterexample reduction algorithm\label{fig:reduction}}
\end{figure}

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes a counterexample \ttp{cex} and the property \ttp{prop}, which
for our purposes, we can suppose has the type \ttp{a -> Bool}.  The reduction
begins at the first sub-value of \ttp{cex}.  Why not begin with \ttp{cex}
itself?  The idea is that

\todo{optimization: check value directly}

%% For example, consider the following types that give us a binary tree
%% %
%% \begin{code}
%% data T = N Int
%%        | B T T
%% \end{code}
%% %
%% \noindent
%% and suppose that the property is that the sum of the leaves is less than 10.
%% Starting from the following counterexample
%% %
%% \lstset{caption={A counterexample value.}
%%        ,label=lst:val}
%% \begin{lstlisting}
%% B (B (N 2) (N 3)) 
%%   (B (N 4) (B (N 5) (N 6)))
%% \end{lstlisting}
%% %
%% \noindent
%% Consider the sub-value \ttp{B (N 4) (B (N 5) (N 6))}.  Some example values of
%% structurally smaller well-typed values include the following:
%% %
%% \begin{code}
%% N 2
%% B (N 8) (N 7)
%% B (N 0) (N 1)
%% \end{code}
%% %
%% \noindent
%% Replacing the sub-value, we get the following new values, each of which is
%% smaller than the value in Listing~\ref{lst:val}:
%% %
%% \begin{code}
%% B (B (N 2) (N 3)) 
%%   (N 2)

%% B (B (N 2) (N 3)) 
%%   (B (N 8) (N 7))

%% B (B (N 2) (N 3)) 
%%   (B (N 0) (N 1))
%% \end{code}

In this work, we limit ourselves to algebraic data types.

\todo{GADTs?}
